Hopf Algebras\vspace{-1ex}

### 2019-03-02 Hopf Algebra Notes/Hopf Algebras.tex
Updates

\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\gldim}{gldim}
\DeclareMathOperator{\projdim}{projdim}
\DeclareMathOperator{\findim}{findim}

@@

@@

%Hopf algebra stuff
\newcommand*{\Vectk}{\operatorname{Vect}_k}
\newcommand*{\Algk}{\operatorname{Alg}_k}
\newcommand*{\Coalgk}{\operatorname{Coalg}_k}
\newcommand*{\calM}{\mathcal{M}}
\DeclareMathOperator{\Coalg}{Coalg}
\DeclareMathOperator{\Alg}{Alg}
\newcommand*{\scrT}{\ensuremath\mathscr{T}}
\newcommand*{\Ch}{\mathbf{Ch}}

@@

Some more results (mostly for finite dimensional algebras):
\begin{thm}[Radford `75]
	If $V$ is finite dimensional, then the antipode $S$ has finite order:
	\[S^d=\id_V\]
\end{thm}
\begin{thm}[Larson-Radford `88]
	Suppose $\ch k\ne 0$ (can be relaxed but gets more ugly). Then the following are equivalent:
	\begin{enumerate}
		\item $\gldim V=0$ ($V$ is semisimple Artinian)
		\item $\gldim V^*=0$
		\item $S^2=\id_V$
	\end{enumerate}
\end{thm}
\begin{thm}[Nichols-Zoeller `89]
	Let $V$ be finite dimensional. If $Q$ is a Hopf subalgebra of $V$, then $_WV$ and $V_W$ are free.
\end{thm}

@@

\subsection{Looking Forward}
On Friday, we may see the following result:
\begin{thm}
	Under some mild hypothesis (including finite dimensionality),
	\[\projdim_V k=\gldim V\]
\end{thm}
\end{document}

>>>>

\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\gldim}{gldim}
\DeclareMathOperator{\projdim}{projdim}
\DeclareMathOperator{\injdim}{injdim}
\DeclareMathOperator{\findim}{findim}

@@

\DeclareMathOperator{\End}{End}

@@

%Hopf algebra stuff
\newcommand*{\Vectk}{\operatorname{Vect}_k}
\newcommand*{\Algk}{\operatorname{Alg}_k}
\newcommand*{\Coalgk}{\operatorname{Coalg}_k}
\newcommand*{\calM}{\mathcal{M}}
\DeclareMathOperator{\Coalg}{Coalg}
\DeclareMathOperator{\Alg}{Alg}
\newcommand*{\scrT}{\ensuremath\mathscr{T}}
\newcommand*{\Ch}{\mathbf{Ch}}
\newcommand*{\g}{\mathfrak{g}}
\DeclareMathOperator{\Der}{Der}

@@

Some more results (mostly for finite dimensional algebras):
\begin{thm}[Radford `75]
	If $V$ is finite dimensional, then the antipode $S$ has finite order:
	\[S^d=\id_V\]
\end{thm}
\begin{thm}[Larson-Radford `88]\label{thm-LR88}
	Suppose $\ch k\ne 0$ (can be relaxed but gets more ugly). Then the following are equivalent:
	\begin{enumerate}
		\item $\gldim V=0$ ($V$ is semisimple Artinian)
		\item $\gldim V^*=0$
		\item $S^2=\id_V$
	\end{enumerate}
\end{thm}
\begin{thm}[Nichols-Zoeller `89]
	Let $V$ be finite dimensional. If $Q$ is a Hopf subalgebra of $V$, then $_WV$ and $V_W$ are free.
\end{thm}

@@

\subsection{Looking Forward}
On Friday, we may see the following result:
\begin{thm}
	Under some mild hypothesis (including finite dimensionality),
	\[\projdim_V k=\gldim V\]
\end{thm}

\section{February 13, 2019}
Here's the theorem for today (with the hypotheses in place):
\begin{thm}\label{thm-213}
	Assume $\ch k=0$ and $S^2=\id_V$. Then 
	\[\gldim V=\projdim{_VM}\]
	for all finite dimensional $V$ modules $M$.
\end{thm}
Another result is the theorem~\ref{thm-LR88}:
\begin{prf}
	Assuming theorem~\ref{thm-213}, (c)$\Rightarrow$(a) is obvious. That's only part of it, 
	obviously. :)
\end{prf}
\subsection{Example of the Day}
\begin{ex}
	$\mathcal{U}(\g)$, the universal enveloping algebra which (as an algebra) is isomorphic to
	$k\langle \g\rangle/(xy-yx-[xy])$. As a coalgebra, we define $\Delta(x)=x\otimes 1+1\otimes x$,
	$\varepsilon=0$ and $S(x)=-x$.

	This is a nice example for today, because notice that $S^2(x)=x$ and in fact $S^2=\id_{\mathcal{U}(\g)}$,
	so we get that $\gldim\mathcal{U}(\g)=\projdim_{\mathcal{U}(\g)}M$ for any finite dimensional $M$ (by 
	the theorem).
\end{ex}
\subsection{Today's Content}
\begin{lem}[Reflection Principle]\label{lem-ref]}
	Let $A$ be an algebra. Then
	\begin{enumerate}
		\item $(-)^*=\Hom)k(-,k):({_A\calM})_{fd}\to(\calM_A)_{fd}$ is a contravariant equivalence.
		\item $(M^*)^*\cong M$ for each $M\in{_AM}_{fd}$. In linear algebra, this says that we have a perfect pairing $M\otimes M^*\to k$.
	\end{enumerate}
\end{lem}
\begin{lem}[Restriction of Scalars]\label{lem-res1]}
	Let $f:A\to B$ be a map of algebras. Then
	\begin{enumerate}
		\item There is a functor $f^*:{_B\calM}\to{_A\calM}$; also 
		\item $f^*:\calM_B\to\calM_A$.
	\end{enumerate}
\end{lem}
\begin{lem}[Restriction of Scalars]\label{lem-res2}
	Let $f:A\to B$ be an antihomomophism of algebras. Then
	\begin{enumerate}
		\item There is a functor $f^*:{_B\calM}\to{\calM_A}$; also 
		\item $f^*:\calM_B\to{_A\calM}$.
	\end{enumerate}
\end{lem}
\begin{lem}
	Let $V$ be a Hopf algebra with antipode $S$. Then 
	\[\tilde S\circ (-)^*:{_V\calM}\to \calM_V\to {_V\calM}\]
	is a contravariant functor.
\end{lem}
\begin{cor}
	If $V$ is finite dimensional, then $\tilde S\circ(-)^*$ is a duality of $(_V\calM)_{fd}$.
\end{cor}
For the future we will use the notation $M^{*S}=\tilde S(M^*)$. Notice that for all $M\in{_V\calM},$ $M^{*S}\in{_V\calM}$.

This leads us to our proposition:
\begin{thm}
	$V$ is a Hopf algebra.
	\begin{enumerate}
		\item If $M\in{_V\calM}$ then there is a natural morphism
		\[ev_M:M^{*S}\otimes M\to {_Vk}\]
		\item if $M\in{_V\calM_{fd}}$, then there is a natural map
		\[coev_M:{_Vk}\to M\otimes M^{*S}\]
	\end{enumerate}
\end{thm}
\begin{prf}
	$ev_M(f\otimes m)=f(m)\in k$ is the map we're talking about. For all $v\in V$, 
	\begin{align*}
		ev_M(v\cdot(f\otimes m))&=ev_M(v_1f\otimes v_2 m)\\
		&=ev_M(fS(v_1)\otimes v_2m)\\
		fS(v_1)(v_2 m)\\
		&= f(S(v_1)v_2m)\\
		&= f(\varepsilon(v) m)=\varepsilon(v)f(m)=v\cdot ev_M(f\otimes m).
	\end{align*}

	If instead $M$ is finite dimensional, pick a $k$ basis $m_1,\dots,m_d$ for $M$ and let 
	$m_i^*$ be the dual basis. Set $coev_M(1)=\sum m_i\otimes m_i^*=\Phi$. Then use the fact that (under the identification of 
	$M\otimes M^*$ with $\Hom_k(M,M)$) we get $\Phi=\id_M$. Then prove 
	\[coev_M(v\cdot 1)=v\cdot coev_M(1).\]
\end{prf}
\begin{defn}
	The map $ev_M$ above is called the \textbf{evaluation map} and $coev_M$ is called the 
	\textbf{coevaluation map.}
\end{defn}

\begin{lem}
	\begin{enumerate}
		\item $(M^{*S})^{*S}\cong \tilde S^2(M)$ for all finite dimensional $M\in{_V\calM}$.
		\item If $S^2=\id$ then $(M^{*S})^{*S}\cong M$.
	\end{enumerate}
\end{lem}
\begin{lem}
	Let $M\in {_VM}_{fd}$ and suppose $S^2=1$. Then 
	\[ev_{M^{*S}}\circ coev_M=(\dim M)\id_k\]
\end{lem}

\section{February 15, 2019}

Here we use the (slightly extended) lemma:
\begin{lem}
	\begin{enumerate}
		\item $ev_M\circ coev_{M^{*S}}=(\dim M)\id_k$
		\item $\frac{1}{\dim M}ev_M\circ coev_{M^{*S}}=\id$
		\item (since the identity factors through $M^{*S}\otimes M$) $M^{*S}\otimes M\cong {_Vk}\oplus N$ as $V$-modules.
		\item $\projdim k\le \projdim M^{*S}\otimes M$.
	\end{enumerate}
\end{lem}
Then using this lemma, we can prove the theorem~\ref{thm-213}:
\begin{prf}
	Basically use this, Lorenz-Lorenz, and the result bounding $\projdim$ of a tensor module by its factors.
\end{prf}

Notice that if we change things up slightly we get a new theorem:
\begin{thm-prime}{thm-213}
	Let $V$ be aHopf algebra. Suppose that $S^2=\id$. Then 
	\[\projdim_V M=\gldim V\]
	for all $M\in {_VM}_{fd}$ such that $\dim M\ne 0$ in $k$.
\end{thm-prime}

\begin{conj}
	Let $V$ be a Noetherian Hopf algebra over a field $k$ of characteristic zero and $S^2=\id$.
	Then
	\[\gldim V <\infty\]
\end{conj}

\begin{cor}
	If $V$ is a finite module over $\mathcal Z(V)$, its center, then the above conjecture holds.
\end{cor}

\subsection{Today's example}
He did the quantized enveloping algebra of $\sl_2$! Might be worth looking up.

An interesting part of this example is that $S^2$ is \textbf{not} the identity! In fact, 
this example is important in constructing Joon (sp?) polynomials in knot theory.

\subsection{Today's Homework}
\begin{prob}
	Prove the following ``dimensional shift'' lemma: Let
	\[0\to X\to P\to Y\to 0\]
	be a SES of $A$ modules, where $P$ is projective. If $M$ is an $A$ module, then for all $n\ge 1$:
	\[\Ext_A^{n+1}(Y,M)\cong\Ext^{n}_A(X,M).\]
\end{prob}
\begin{sol}
	Take a projective resolution for $X$ and try to attach it to the SES.
\end{sol}
\subsection{Classifying Algebras}
We have the (descending) chain of classes of algebras:
\begin{itemize}
	\item Algebras
	\item Noetherian Algebras
	\item Cohen-Macaulay Algebras
	\item Gorenstein Algebras
	\item Complete Intersections
	\item Hypersurface Rings
	\item Regular Algebras
\end{itemize}
The question we have is: where to (Noetherian) Hopf algebras lie in this hierarchy?

If we restrict to finite-dimensional Hopf algebras, Brown-Goodearl proved that Noetherian Hopf algebras
are all Gorenstein rings.

\begin{defn}
	A Noetherian algebra $A$ is called \textbf{regular} if $\gldim A<\infty$.
\end{defn}
\begin{rmk}
	Notice that since $A$ is Noetherian, we didn't have to talk about left vs. right global dimension.
\end{rmk}
\begin{ex}
	If $L$ is a finite dimensional Lie algebra, then $U(L)$ is regular:
	\[\gldim U(L)=\dim L.\]
\end{ex}
\begin{defn}
	An element $x\in A$ is called \textbf{normal} if $xA=Ax$ (NOT $xa=ax$)
\end{defn}
\begin{defn}
	A Noetherian algebra $A$ is called a \textbf{hypersurface ring} if there exists a Noetherian 
	regular algebra $B$ and a normal non-zero-divisor $x\in B$ such that $A\cong B/(x)$
\end{defn}
\begin{ex}
	$\ch k=p>0$, $V=k[x]/(x^p)$, a Hopf algebra with comultiplication $\Delta(x)=x\otimes 1+1\otimes x$.
	Then $V$ is a hypersurface ring.
\end{ex}
\begin{defn}
	A Noetherian algebra $A$ is called a \textbf{complete intersection} if there is a Noetherian regular algebra
	$B$ and a sequence of elements $x_1,\dots,x_d$ such that:
	\begin{itemize}
		\item $x_1$ is a non-zero-divisor in $B$.
		\item $x_i$ is a NZD in $B/(x_1,\dots,x_{i-1})$ for all $i$
		\item $A\cong B/(x_i)$.
	\end{itemize}
\end{defn}

\begin{defn}
	A Noetherian algebra $A$ is called \textbf{Gorenstein} if 
	\[\injdim{_AA}=\injdim A_A<\infty.\]
\end{defn}
\begin{defn}
	A Noetherian algebra $A$ is called \textbf{Cohen-Macaulay} if there is an $A$-bimodule
	$M$ such that
	\begin{itemize}
		\item $\injdim {_AM}<\infty$, $\injdim M_A<\infty$
		\item $\End(M_A)\cong A,$ $\End({_AM})\cong A^{op}$
		\item ${_AM}$ and $M_A$ are Noetherian.
	\end{itemize}
\end{defn}

\section{February 20, 2019}
Last time we wrote out a big chain of inclusions for algebras. We didn't really discuss why all complete
intersections are Gorenstein algebras. Actually, this is one of the harder inclusions to prove. We will at least 
try today, but will definitely get to it by Friday.

\subsection{Example of the day}
\begin{defn}
	$V=k[g^{\pm 1}][y;\delta]$, an Orr extension of the Laurent polynomial ring by the derivation $\delta=(g^n-g)\frac{d}{dg}$.
	For those (me) wo haven't seen Orr extensions before, we also have
	\[V\cong k\langle g, g^{-1}, y\rangle/(yg-gy-(g^n-g), gg^{-1}-1, g^{-1}g-1)\]

	This is a domain where $\gldim V=2$ (also the GK dimension, which we haven't defined). As a coalgebra, we get
	$g$ is grouplike and $\Delta y= y\otimes g^{n-1}+1\otimes y$, $\varepsilon(y)=0$, and $S(y)=-yg^{-(n+1)}$.

	Notice that $V$ is regular.
\end{defn}
\subsection{Problem of the day}
\begin{prob}
	Let $x$ be a normal non-zero-divisor in $A$ where $xA=Ax$. Then for all $a\in A$ let $\sigma\in\Aut A$ be defined such that
	$\sigma(a) x= xa$.

	Let $M\in {_A\calM}$ and define a left $A$ action on ${^\sigma M}$ whose underlying set is $M$ and the action is defined by $a\cdot m=\sigma(a)m$. In terms of the action (morphism), this is 
	equivalent to precomposing with $\sigma:$ 
	\[A\xrightarrow{\sigma} A\xrightarrow{\rho}\End(M)\]

	Prove the following lemma:
	\begin{lem}[Rees Lemma]
		Let $x$ be a normal non-zero-divisor in $A$. If $L$ is a left $A/(x)$-module and $M$ is an $x$-torsion-free $A$-module, then
		\[\Ext_{A/(x)}^n(L,M/xM)\cong \Ext_A^{n+1}(L,{^\sigma M}).\]
	\end{lem}
\end{prob}

\subsection{Today's Results}
We begin with some corollaries of Rees Lemma:
\begin{cor}
	\begin{enumerate}
		\item $\Ext_{A/(x)}^n(L, A/(x))\cong \Ext_A^{n+1}(L,{^\sigma A})\cong\Ext_A^{n+1}(L,A)$
		\item $\injdim_{A/(x)}(A/(x))\le \injdim_AA-1$
		\item If $A$ is Gorenstein, so is $A/(x)$.
	\end{enumerate}
\end{cor}
\begin{rmk}
	Then the idea here is that we just have to prove that regular rings are Gorenstein, since then complete intersections
	are just (finite) sequences of quotients by regular elements.
\end{rmk}
\subsection{Doing $\Ext$ again}
Let $P^\bullet$ and $Q^\bullet$ be complexes of $A$-modules. Then $\Hom_A(P^\bullet, Q^\bullet)$ is a complex. Define
\[\Hom_A(P^\bullet, Q^\bullet):=\prod_{i\in\Z}\Hom_A(P^i,Q^{i+n}=\prod_{-i+j=n}\Hom(P^i,Q^j)\]
where we define the boundary map to be 
\[d_{\Hom_A(P^\bullet,Q^\bullet)}:(f_i)_{i\in\Z}\mapsto (d_Q^{n+1}\circ f^i - (-1)^nf^{i+1}\circ d_P^i)_{i\in\Z}\]
\begin{prob}
	The \textbf{real} homework for this class (the other one is what we're showing, apparently) is to show that this is a
	chain complex -- i.e. that $d^2=0$.
\end{prob}

Define the projective resolution to be the regular projective resolution, but do not include $M$ itself. But then we have
\begin{align*}
	R\underline{\Hom}=\Ext_A^n(M,N)&:=H^n(\Hom_A(P^\bullet_M,N))\\
	&:=H^n(\Hom_A(P_M^\bullet,P_N^\bullet))\\
	&:=H^n(\Hom_A(I_M^\bullet, I_N^\bullet))\\
	&:=H^n(\Hom_A(M,I_N^\bullet))
\end{align*}

\subsection{Proof plz}
Today we will try to get $n=0$ in the inductive argument.
\begin{lem}
	\begin{enumerate}
		\item  There is a short exact sequence of left $A$ modules
		\[0\to {^\sigma A}\xrightarrow{l_x}A\to A/(x)\to 0\]
		where $l_x$ sis left multiplication by $x$.
		\item If $M$ is $x$-torsion-free, then there is a SES of left $A$ modules
		\[0\to {^\sigma M}\xrightarrow{l_x}M\to M/xM\to 0\]
	\end{enumerate}
\end{lem}
The second item implies the first and neither is hard to show directly.
\begin{lem}
	Let $M$ be an $x$-torsion-free left $A$ modules and let $I_M$ be an injective resolution of $M$. THen there is a morphism of complexes
	$l_x:{^\sigma I_M}\to I_m$.
\end{lem}
Again, not too hard to check.

Notice now that $xL=0$ by the hypotheses. Combining this with the last result, we have
\begin{lem}
	\begin{enumerate}
		\item For any $L\in {_A\calM}$,
		\[\Ext_A^n(L,l_x):\Ext_A^n(L,{^\sigma M})\to \Ext_A^n(L,M)\]
		\item If $xL=0$ then the above map is zero for all $n\ge 0$.
	\end{enumerate}
\end{lem}
Again, follow your nose a bit. $f:L\to {^\sigma I^i}$ is an $A$ homomorphism and pull that $x$ in.

Applying $\Hom_A(L,-)$ to $0\to {^\sigma M}\xrightarrow{l_x}M\to M/xM\to 0$, we obtain the long exact sequence
for $\Ext$. But then notice that $\Hom_A(L,M)=\Ext_A^1(L,M)=0$ since $M$ is $x$-torsion-free (and $L$ is killed by $x$).
and since $\Ext_A^1(L,{^\sigma M})\cong \Hom_A(L,M/xM)\cong\Hom_{A/(x)}(L,M/xM)$, we get the base case proven.

\section{February 22, 2019}
Today we are going to finish the proof of Rees, then give some remarks, then give the example of the day. :)

\subsubsection{Continuation of Proof of Rees}
Last time we proved that $\Ext_{A}^1(L,{^\sigma M})\cong \Ext_{A/(x)}^0(L,M/xM)$. Now we proceed by induction.

Consider the SES $0\to K\to F\to L\to 0$ of left $A/(x)$-modules where $F$ is free. Then we get a long exact sequence
\begin{align*}
	0\to \Hom_{A/(x)}(L,M/xM)\to \Hom_{A/(x)}&(F,M/xM)\to \Hom_{A/(x)}(K,M/xM)\\
	&\to \Ext^1_{A/(x)}(L,M/xM)\to\Ext^1_{A/(x)}(F,M/xM)\to\cdots
\end{align*}
as well as the sequence
\begin{align*}
	\cdots \to \Ext^1_{A}(L,{^\sigma M})\to \Ext^1_{A}&(F,{^\sigma M})\to \Ext^1_{A}(K,{^\sigma M})\\
	&\to \Ext^2_{A}(L,{^\sigma M})\to \Ext^2_{A}(F,{^\sigma M}\to\cdots)
\end{align*}
and we have isomorphisms between all the corresponding parts (except the fourth term). THe first three 
follow from $n=0$ and the last one follows since $F$ is a projective $A/(x)$ module and since $F$ has projective dimension (over $A$) 
less than or equal to one. Then five lemma proves that this holds for $n=1$.

Then you can just extend both sequences and iteratively prove this holds for each term. Woot. :)

\subsection{Return to Classifications}
Recall we have the (descending) category chain
\begin{itemize}
	\item Algebras
	\item Noetherian Algebras
	\item Cohen-Macaulay Algebras
	\item Gorenstein Algebras
	\item Complete Intersections
	\item Hypersurface Rings
	\item Regular Algebras
\end{itemize}
Now we can show by applying Rees directly:
\begin{cor}
	$\injdim A/(x)\le \injdim A$
\end{cor}
Which then allows us to prove:
\begin{cor}
	Every complete intersection ring is Gorenstein.
\end{cor}
\begin{prf}
	Let $A$ be a CI ring. Then there exists a Noetherian regular ring $B$ and a sequence of elements $b_1,\dots, b_n\in B$ 
	such that each $b_i$ is a normal nonzerodivisor on $B/(b_1,\dots,b_{i-1})$.

	Since $B$ is regular, it is Gorenstein. By the corollary $B/(b_1)$ is Gorenstein. Continuing this way, $A$
	itself is Gorenstein and we are done. 
\end{prf}

\subsection{Where do Hopf algebras fit in?}
Recall that we proved that (finite dimensional) Hopf algebras are Frobenius algebras. Well it ends up
that Frobenius algebras are also Gorenstein! Thus all (f.d) Hopf algebra is Gorenstein.

Recall the Brown-Goodearl conjecture: all Noetherian Hopf algebras are Gorenstein. This is still an open conjecture!
But it gets worse!
\begin{conj}
	Every Noetherian Hopf algebra is a complete intersection.
\end{conj}
\begin{conj}
	Every finite dimensional Hopf algebra is a complete intersection.
\end{conj}
There are no known counterexamples to either! This has been open for decades with no satisfactory result.
\subsubsection{Even more classes of algebras!}
Inside the category of Gorenstein rings, there are a couple more classes (not related to complete intersections):
We get the decreasing chain
\begin{itemize}
	\item Gorenstein rings
	\item Artin-Schelter Gorenstein
	\item A-S Gorenstein with skew Calabi-Yau property
\end{itemize}
and over at regular algebras we get the decreasing chain
\begin{itemize}
	\item Regular algebras
	\item A-S regular algebras
	\item Skew C-Y algebras
	\item C-Y algebras
\end{itemize}
\begin{defn}
	A Noetherian Gorenstein ring $A$ of $\injdim$ d is called $A-S$ Gorenstein if, for all $S\in{_A\calM}_{fd},$
	\[\Ext_A^i(S,A)=\left\{\begin{array}{lr}0 & i\ne d,\\ \in {_A\cal M}_{fd}, & i=d\end{array}\right.\]
	where the above property is called the Artin-Schelter (AS) property.
\end{defn}
\begin{defn}
	A Noetherian A-S Gorenstein algebra $A$ of injective dimension $d$ is called \textbf{skew Calabi-Yau} if
	\[\Ext_{A^e}^i(A,A^e)=\left\{\begin{array}{lr}0 & i\ne d,\\ {^\sigma A}, & i=d\end{array}\right.\]
	(remember that $A^e\cong A\otimes A^{op}$).
\end{defn}
\begin{defn}
	$A$ is Calabi-Yau if it is skew C-Y and also $\Ext_{A^e}^d(A,A^e)=A$.
\end{defn}
\begin{rmk}
	The idea here (and honestly I missed some parts of the discussion) is that C-Y algebras are ones where we can get 
	Poincar\'e duality.
\end{rmk}
\subsection{Example of the Day}
\begin{ex}
	$V$ is generated by $g,g^{-1},x_2,x_3$ subject to
	\[gg^{-1}=g^{-1}g=1,\quad gx_i=\eta^ix_i g,\quad x_2^3=x_3^2\quad x_2x_3=x_3x_2\]
	where $\eta$ is a primitive $6^{th}$ root of unity.

	Then $\gldim V=\infty$ and it is a kypersurface ring, since
	\[V\cong k[x_1,x_2][g^{\pm 1};\phi]/(x_2^3-x_3^2)\]
	where $\phi:x_i\to \eta^ix_i$.

	As a coalgebra, $g$ is grouplike and $x_i$ are such that $\Delta(x_i)=x_i\otimes 1+g^i\otimes x_i$
	and $\varepsilon(x_i)=0$. Then we get $S(x_i)=-g^{-i}x_i.$
\end{ex}
\subsection{HW for the day}
\begin{prob}
	Let $A$ and $B$ be rings. Let $_AM$, $_AN_B$, and $_BC$. Then 
	\[\Hom_A(N\otimes_B C,M)\cong \Hom)B(C, \Hom_A(N,M))\]
	and if instead $C_A$, $_AN_B$ and $M_B$, then you can do things over $A^{op}$ and $B^{op}$.
\end{prob}

\section{February 25, 2019}
Topic for today: \textbf{Hopf actions on algebras.} One can naturally extend everything we talk about
today to (co)actions on (co)algebras. :)
\subsection{Example of the Day}
\begin{ex}\label{ex-22519}
	$V=k\langle g,g^{-1},x\rangle/(gg^{-1}=1=g^{-1}g)$. Then $g$ is grouplike and $\Delta(x)=x\otimes1+g\otimes x$, 
	$\varepsilon(x)=0$, and $S(x)=-g^{-1}x$.

	We can see that $\gldim V=1$ (use resolution $0\to V\oplus V\to V\to {_V k}\to 0$). Note that
	it is NOT AS-Gorenstein. $\Ext_V^1(k,V)$ is infinite dimensional over $k$ (not easy!).
\end{ex}
\begin{rmk}
	We've seen elements like $x$ above several times -- these elements are called $(1,g)$-primitive 
	(generalizing the concept of primitive, obviously).

	Also, these are examples of \textbf{pointed Hopf algebras}(!!) These are $V$ such that (as a
	coalgebra) the coradical is a sum of one-dimensional modules. This is actually related to the fact 
	that every simple submodule of $V$ is of the form $kg^n$ (in this case $n=1$).
\end{rmk}
\subsection{HW of the Day}
\begin{prob}
	This problem considers (one of the) internal $\Hom$(s) in $_V\calM$. Suppose $V$ has bijective 
	antipode. Then for any $M$ and $N$,
	\[\Hom^{S^{-1}}(M,N):=\Hom_k(M,N)\]
	where a $V$ action is defined by (for $v\in V$ and $f\in\Hom^{S^{-1}}(M,N)$) 
	\[(v\cdot f)(m)=\sum v_2f(S^{-1}(v_1)m)\]
	where, as usual, $\Delta v=\sum v_1\otimes v_2$.
	\begin{enumerate}
		\item Show that this makes $\Hom^{S^{-1}}(M,N)$ into a left $V$-module.
		\item Show that $\Hom^{S^{-1}}(-,-)$ is right adjoint to $-\otimes_k -$.
	\end{enumerate}
\end{prob}
\subsection{Automorphisms and Group Actions}
Let $A$ be an algebra. $\Aut_{alg}(A)$ denotes the \textbf{group of autormorphisms of $A$.}
\begin{defn}\label{def-action}
	Let $G$ be a group. We say $G$ acts on $A$ ($G\curvearrowright A$) fi there is a group homomorphism
	$G\to \Aut_{alg}(A)$.
\end{defn}
\begin{defn}
	Alternatively, we say $G\curvearrowright A$ if
	\begin{itemize}
		\item $A$ is a $G$-module.
		\item $\forall g\in G$, $g(1_A)=1_A$
		\item $\forall a,b\in A$, $g(ab)=g(a)g(b)$
		\item $g$ is bijective.
	\end{itemize}
\end{defn}
\begin{defn}
	Suppse $G\curvearrowright A$. Then the \textbf{invariant subring} of this action is 
	\[A^G:=\{a\in A|g(a)=a,\forall g\in G\}\subseteq A\]
\end{defn}
This was fairly popular around 1880-1960 or so.
\begin{ex}
	$A=k[x_1,x_2]$, $G=\langle\sigma\rangle\cong C_2$ where
	$\sigma\cdot f(x_i)=f(-x_i)$. One can compute that this just switches signs on monomial 
	of odd degree (and fixes the even ones) so
	\[A^G=\left\{\sum c_{ij}x_1^ix_2^j|i+j=2k\right\}\cong k[a,b,c]/(ac-b^2)\]
\end{ex}
Generalizing this, many of the questions of this time revolved around asking what category of 
algebra $k[x_1,\dots,x_n]^G$ is for arbitrary $G$. We may return to this later.

\subsection{Derivations}
\begin{defn}
	Let $A$ be an algebra. A map $d\in\Hom_k(A,A)$ is called a \textbf{derivation} if
	it satisfies the Leibniz rule.
\end{defn}
\begin{lem}
	The collection of all derivations of an algebra from a Lie algebra -- denoted $\Der(A)$.
\end{lem}
\begin{defn}
	Let $L$ be a Lie algebra. We say $L$ acts on $A$ if there is a Lie algebra morphism $L\to \Der(A)$.
	Equivalently, if
	\begin{enumerate}
		\item $A$ is a left $L$-module
		\item $\forall l\in L$, $l(1_A)=0$
		\item $\forall a,b\in A$, $l(ab)=l(a)b+al(b)$.
	\end{enumerate}
\end{defn}
\begin{rmk}
	The second item above is implied by the first (in fact, James included a similar statement in 
	the definition of a derivation) but apparently there is a reason to include it!
\end{rmk}
\begin{ex}
	$A=k[x_1,x_2]$, $d=\frac{\partial}{\partial x_1}$. Then $L=kd$ is a Lie algebra that acts on $A$.

	Then $A^L$ (see below) is $k[x_2]$
\end{ex}
\begin{defn}
	If $L$ is a Lie algebra acting on $A$, then the invariant subring under the action is
	\[A^L=\{a\in A|l(a)=0,\forall l\in L\}\]
\end{defn}

\subsection{Putting These Things Together}
Let $A$ be an algebra and $V$ be a Hopf algebra.
\begin{defn}
	We say $V$ actions on $A$ if
	\begin{itemize}
		\item $A$ is a left $V$-module
		\item $A$ is an algebra object in $({_V\cal M},\otimes)$
	\end{itemize}
	Note that the second item above implies the first!

	Equivalently, $V$ acts on $A$ if
	\begin{itemize}
		\item $A\in {_V\cal M}$
		\item For all $v\in V$, $v\cdot(1_A)=\varepsilon(v)1_A$
		\item For all $a,b\in A$, $v(ab)=\sum v_1(a)v_2(b)$.
	\end{itemize}
\end{defn}

Returning to the example from the beginning of class (\ref{ex-22519}), if $V$ acts on $A$,
then 
\begin{itemize}
	\item $g\in \Aut_{alg}(A)$
	\item $x\in\Der_g(A)$
\end{itemize}
In particular, notice that $x(1)=0$ and $x(ab)=x(a)b+g(a)x(b)$.
\end{document}
